Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
lots of small text improvements

The papers submitted to this panel, and several others, include lots of general
recommendations for processes, practices, tools, etc. One of the papers
suggested that a ``Software Sustainability Institute'' should be vested with
the authority to develop standardized quality processes, a central common
repository, central resources for services and consulting, a think tank of
sorts, and a software orphanage center. The idea of one common repository received
some resistance, with so many compelling alternatives available, e.g. BitBucket or
GitHub. The point for centralization of communication/point of contact was seen as
reasonable, with the statement that ``vested with authority'' is perhaps too strong,
but ``providing tools if needed'' might be more appropriate.

@@

For the purpose of studying software case studies, we classify the software
projects discussed in the workshop in three broad categories. The first is 
`utility software', software tools that enable and/or facilitate the
development of other tools and techniques to carry out scientific work. The second is
`scientific software', software that was developed with an aim to solve a
class of scientific problems. The third is `infrastructure software', software
developed to efficiently utilize new research infrastructure.

These categories have some overlap, and in some cases, the
case studies do not cleanly fit into a single category.
Our classification is based on two criteria: first, the character
of a software project that dominates and where it `seems' to be the best fit, and, second,
the the workshop discussion session where the paper about
the project was discussed.

@@

\subsection{Utility Software}
Utility software potentially has the broadest community impact of the three
categories. This is especially relevant in the modern era of social computing.
In the context of this paper, a utility software project could be defined as
general purpose software applicable to one or many generic tasks and/or
enabling other software to run by providing a suitable environment. Examples
are collaborative development framework such as \emph{GitHub} and
\emph{Bitbucket}, workflow and scripting frameworks such as \emph{Galaxy},
\emph{Swift}, \emph{Globus} and \emph{VisTrails}, and visualization frameworks
such as \emph{visIT}. Often projects start with specific, small set of
requirements and users in mind but expand to provide more generic functionality
and appeal to a wider user community. Owing to its generic nature, utility
software finds a wider users communities which in turn leads to a higher
participation in development. Consequently, the process becomes user-driven and
self sustaining. Software developed in such scenario often benefits from
community driven development, deployment and maintenance best practices. For
instance the \emph{Galaxy} project has significantly benefited from user
community participation in its development. It follows agile software
development practices and implements standard practices such as test-driven
development and socialized bug managing practices via \emph{trello}. Galaxy
\emph{histories} and \emph{toolshed} offers easy community sharing of data and
tools further promoting a collaborative environment. The project very closely
follows the guidelines described in~\cite{Carver_WSSSPE} and many
from~\cite{Prlic_WSSSPE}.

Another paradigm in utility software is the software delivered as service over
the web. With increasing popularity of cloud-based computational environments,
many users are leaning towards tools used as service. \emph{GitHub} can be
arguably considered one such tool, catering to collaborative development. For
scientific users \emph{Globus} based tools are case of service based utility
software discussed during the workshop. The data movement, authentication and
sharing services offered by Globus can be easily used over the web by
collaborating researcher.

\subsection{Scientific software}
Scientific software consists primarily of special purpose software that was developed for a target use-case
scenario or keeping some fixed/frozen requirements in mind. Software projects
pertaining to specific scientific domain often tend to be in a niche and the
user community tends to be small to medium. Numerical accuracy and algorithmic
optimization are some of the paramount requirements of most scientific
software. Advantages of a smaller codebase and requirements include stability,
ease of installation, configurability and integration into larger systems.
While the software can stay stable and require relatively low maintenance, the
responsibility is often on the shoulders of a very few developers. Community
participation in the development tends to be low. Development is linear and
simplistic with a limited scope to follow software best practices. Domain
science oriented software developed to solve  specific problems, e.g.,
\emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. Sustainability
of such software is often a significant challenge. Many submissions reported the
software considered a `byproduct' of the actual research.  Others contended that
the software was not the main funded part of their research. One exception was
\emph{Kitware}, which while being a software product specializing in scientific
process, has a core focus of developing communities around software processes.
One example of this process is the development of the \emph{CMake} build
utility, which started out as a building tool for \emph{ITK} but grew to become
a generic build utility for C++ projects.

>>>>

The papers submitted to this panel, and several others, include lots of general
recommendations for processes, practices, tools, etc. One of the papers
suggested that a ``Software Sustainability Institute'' should be vested with
the authority to develop standardized quality processes, a central common
repository, central resources for services and consulting, a think tank of
sorts, and a software orphanage center. The idea of one common repository received
some resistance, with so many compelling alternatives available, e.g. Bitbucket or
GitHub. The point for centralization of communication/point of contact was seen as
reasonable, with the statement that ``vested with authority'' is perhaps too strong,
but ``providing tools if needed'' might be more appropriate.

@@

\note{We should use either single or double quotations consistently throughout the paper.}
For the purpose of studying software case studies, we classify the software
projects discussed in the workshop in three broad categories. The first is 
`utility software', software tools that enable and/or facilitate the
development of other tools and techniques to carry out scientific work. The second is
`scientific software', software that was developed with an aim to solve a
class of scientific problems. The third is `infrastructure software', software
developed to efficiently utilize new research infrastructure.

These categories have some overlap, showing in some cases where
case studies do not cleanly fit into a single category.
Our classification is based on two criteria: first, the character
of a software project that dominates and where it `seems' to be the best fit, and, second,
the the workshop discussion session where the paper about
the project was discussed.

@@

\subsection{Utility Software}
Utility software potentially has the broadest community impact of the three
categories. This is especially relevant in the modern era of social computing.
In the context of this paper, a utility software project could be defined as
general purpose software applicable to one or many generic tasks and/or
enabling other software to run by providing a suitable environment. Examples
are collaborative development frameworks such as \emph{GitHub} and
\emph{Bitbucket}, workflow and scripting frameworks such as \emph{Galaxy},
\emph{Swift}, \emph{Globus} and \emph{VisTrails}, and visualization frameworks
such as \emph{VisIT}. Often projects start with a specific, small set of
requirements and users in mind but expand to provide more generic functionality
and appeal to a wider user community. Owing to its generic nature, utility
software finds a wider user community which in turn leads to a higher
participation in development. Consequently, the process becomes user-driven and
self-sustaining.

Software developed in such a scenario often benefits from
community driven development, deployment and maintenance best practices. One
such example is the \emph{Galaxy} project. It follows agile software
development practices and implements standard practices such as test-driven
development and socialized bug managing practices via \emph{trello}. Galaxy
\emph{histories} and \emph{toolshed} offers easy community sharing of data and
tools further promoting a collaborative environment. The project very closely
follows the guidelines described in~\cite{Carver_WSSSPE} and many
from~\cite{Prlic_WSSSPE}.

Another paradigm in utility software is the software delivered as service over
the web. With increasing popularity of cloud-based computational environments,
many users are leaning towards tools used as service. \emph{GitHub} and
\emph{Bitbucket} can be arguably considered to be such tools, catering to
collaborative development. For scientific users \emph{Globus}-based tools are
a case of service-based utility software discussed during the workshop. The
data movement, authentication and sharing services offered by Globus can be
easily used over the web by collaborating researchers.

\subsection{Scientific software}
Scientific software consists primarily of special-purpose software that was
developed for a target use-case scenario or keeping some fixed/frozen
requirements in mind. Software projects pertaining to specific scientific
domain often tend to be in a niche and the user community tends to be small to
medium. Numerical accuracy and algorithmic optimization are some of the
paramount requirements of most scientific software. Advantages of a smaller
codebase and requirements include stability, ease of installation,
configurability and integration into larger systems. While the software can
stay stable and require relatively low maintenance, the responsibility is often
on the shoulders of a very few developers. Community participation in the
development tends to be low. Development is linear and simplistic with a
limited scope to follow software best practices. Domain science oriented
software is developed to solve  specific problems, e.g., \emph{DUNE},
\emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. Sustainability of such
software is often a significant challenge. Many submissions reported the
software is considered a `byproduct' of the actual research.  Others contended
that the software was not the main funded part of their research. One exception
was \emph{Kitware}, which while being a software product specializing in the
scientific process, has a core focus of developing communities around software
processes. One example of this process is the development of the \emph{CMake}
build utility, which started out as a building tool for \emph{ITK} but grew to
become a generic build utility for C++ projects.
